% !TEX root = ../IS.tex
\chapter{Future Work}
We have seen with a limited survey sample that video game AI models are more challenging for players when they have larger sets of possible actions within the game. However, it would be useful to create a full extensive-tree model of the updated game, like the extensive tree created for the simpler version of the game. With such a tree, deeper analysis can be done on the effectiveness of each AI model. When the outcomes of the game are known, a more complex AI model can be created to specifically seek their winning outcomes. Additionally, this tree could help balance the game. If one player can reliably win using the same strategy each game, then certain parts of the game can be tweaked to mitigate such outcomes; for instance, the amount of damage dealt by a particular action could be decreased, or the probability of missing a strong attack could be increased. Given the random chance of missing strong attacks, there likely is not a simple solution to be found by backwards induction, but using the percentage of missed attacks in a mixed strategy could help to determine the optimal action.\\

Alternative machine learning models can be used to implement the AI. With the complete version of the game, an artificial neural network can be made to simulate the game and learn to play optimally. In some ways, this would be more efficient than attempting to solve the entire game in an extensive-form tree, but the neural network would not show why its decisions were optimal.\\

The balance of healing turns can also be adjusted for better results. Originally, the values of 5 points of damage lost on an attack and 4 points of damage gained from healing were chosen to prevent infinitely long games in Figure \ref{fig:decisionTree1}. However, since later versions of the game limited both players to two healing turns per game, this issue does not occur. If \textit{Cave Escape} changed the point value for healing turns, and the values of a normal attack and healing were tweaked accordingly, then it is likely that healing would become a much more viable strategy, as opposed to the current version where healing tends to be a detriment.